---
title: "Statistical Analysis with rstatix: (Taylor's Version)"
author: "Your Name Here"
date: "12-08-2024"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Introduction

Statistical analysis is a crucial aspect of data science and research. In this guide, we'll explore the functionality of the `rstatix` package, a powerful tool in the R statistical ecosystem. To illustrate its application, we'll use the `taylor` package as our example dataset, focusing on summary statistics, t-tests (independent and paired), correlation plots, and ANOVA.

## Install and Load Packages

Before we begin, ensure that you have the required packages installed:

```{r}
#install.packages("rstatix")
#install.packages("taylor")

library(rstatix)
library(taylor)
library(tidyverse)
```

# Taylor package
The goal of {taylor} is to provide easy access to a curated data set of Taylor Swift songs, including lyrics and audio characteristics. Data comes Genius and the Spotify API.
https://taylor.wjakethompson.com/

```{r}
data("taylor_albums")
data("taylor_all_songs")
```

## Summary Statistics
Summary statistics provide a quick overview of the main characteristics of a dataset. The rstatix package makes it easy to compute these statistics.

```{r}
taylor_all_songs %>% 
  get_summary_stats(duration_ms, danceability, energy, type = "common")
```

This will display key statistics such as mean, standard deviation, and confidence intervals for the a few variables of interest in all songs, durations of the song, dancability score, and energy score.

## Group By Summary

You can do the same thing but group by a variable

```{r}
taylor_all_songs %>%
  group_by(album_name) %>% 
  get_summary_stats(duration_ms, danceability, energy, type = "mean_sd")
```


# T-Tests
T-tests are used to compare means between groups. The rstatix package simplifies the process of conducting t-tests, both independent and paired.

## Independent T-Test
Whats the dancability between evermore and lover album? What about evermore and folklore?
```{r}
# Example: Independent t-test
two_ind <- taylor_all_songs %>% 
  filter(album_name == "evermore" | album_name == "Lover")

ggplot(two_ind, aes(album_name, danceability)) +
  geom_boxplot()

lover_ever <- two_ind %>%
  t_test(danceability ~ album_name, paired = FALSE)
lover_ever


soft_ind <- taylor_all_songs %>% 
  filter(album_name == "evermore" | album_name == "folklore")

ggplot(soft_ind, aes(album_name, danceability)) +
  geom_boxplot()

folk_ever <- soft_ind %>%
  t_test(danceability ~ album_name, paired = FALSE)
folk_ever
```

## Paired T-Test
Now compare the rerecorded albums for taylors version in terms of length of time?

```{r}
## cleaning data to be paired

taylors_version <- taylor_albums %>% 
  filter(str_detect(album_name, "(Taylor's Version)")) %>% 
  pull(album_name)
## 4 taylor albumns
old_version <- str_remove_all(taylors_version, " \\(Taylor's Version\\)")

grouped <- taylor_all_songs %>% 
  filter(album_name %in% c(taylors_version, old_version)) %>%
  mutate(taylor_version = ifelse(str_detect(album_name,"(Taylor's Version)"), 1, 0)) %>% 
  select(duration_ms, track_name, taylor_version, album_name) %>% 
  filter(!str_detect(track_name, "From The Vault|Remix|Acoustic Version|Original Demo Recording|Piano Version")) %>%
  mutate(track_name = str_remove_all(track_name, " \\(Taylor's Version\\)"),
         album_name = str_remove_all(album_name, " \\(Taylor's Version\\)")) %>% 
  pivot_wider(id_cols = c(track_name, album_name), names_from = taylor_version, values_from = duration_ms) %>% 
  filter(!is.na(`0`) & !is.na(`1`)) %>% 
  rowid_to_column(var = "id") %>% 
  pivot_longer(cols = c(`0`,`1`), names_to = "taylors_vers",
               values_to = "duration")

# create plot using ggplot() and geom_boxplot() functions 
ggplot(grouped, aes(taylors_vers, duration, fill=taylors_vers)) + 
geom_boxplot()+ 
# geom_point() is used to make points at data values 
geom_point(alpha = 0.3)+ 
# geom_line() joins the paired datapoints 
geom_line(aes(group=id), alpha = 0.3)


# Example: Paired t-test
t_test_paired_result <- grouped %>%
  t_test(duration ~ taylors_vers, paired = TRUE)
t_test_paired_result
```

## Correlation Plots
Correlation plots are useful for visualizing relationships between variables. The rstatix package facilitates correlation analysis.

```{r}
# Example: Correlation plot
music_scores <- taylor_all_songs %>%
  select(danceability, energy, speechiness, acousticness, liveness)

correlation_matrix <- music_scores %>%
  cor_mat()
correlation_matrix
correlation_matrix %>%
  cor_mark_significant()
# Visualize correlation matrix
correlation_matrix %>%
  cor_plot()

```

This will generate a visual representation of the correlation matrix, making it easier to identify patterns and relationships, like energy and accousticness being negatively correlated.

# Analysis of Variance (ANOVA)
ANOVA is employed to compare means between two or more groups. The rstatix package supports one-way and two-way ANOVA.

## One-Way ANOVA

```{r}
# Example: One-way ANOVA
grouped %>%
  anova_test(duration ~ taylors_vers)

```

The anova_result object will contain the ANOVA table with F-statistics and p-values.

Two-Way ANOVA

```{r}
ggplot(grouped, aes(taylors_vers, duration)) +
  geom_boxplot() +
  geom_line(aes(group = id), alpha = 0.3) +
  facet_wrap(~album_name)

# Example: Two-way ANOVA
grouped %>%
  anova_test(duration ~ taylors_vers + album_name)


```



# Conclusion
The rstatix package streamlines statistical analysis in R, providing a convenient and tidy approach. Through examples with the taylor dataset, we've covered summary statistics, t-tests, correlation plots, and ANOVA. These tools empower researchers and data scientists to derive meaningful insights from their data.

On your own...

Can you run anovas or t-tests on different keys of the song and their dancability or energy? Or speechiness/liveness on if the song is explicit or not?