---
title: "Lab 3"
author: "Your Name Here"
date: "09-29-2024"
output: html_document
---

## Joins

In the tidyverse, joining two datasets together is a way to combine data from different sources based on common variables. The dplyr package within the tidyverse provides functions to perform these joins.

### Inner Join

An inner join combines rows from two datasets where there's a match between the specified variables. Rows with no matching values are excluded. Inner joins return results if the keys are matched in BOTH tables.

    inner_joined_data <- dataset1 %>%
      inner_join(dataset2, by = "common_column")
      
### Practice with Flights Data      

```{r examine_data, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(mdsr)
library(nycflights13)

head(airlines)
head(flights)

```


```{r inner}
#Tidyverse way
flights_joined <- flights %>% 
  inner_join(airlines, by = c("carrier" = "carrier"))
glimpse(flights_joined)
 
flights_joined %>% 
  select(carrier, name, flight, origin, dest) %>% 
  head(3)
```


"In an inner_join(), the result set contains only those rows that have matches in both tables. In this case, all of the rows in flights have exactly one corresponding entry in airlines, so the number of rows in flights_joined is the same as the number of rows in flights (this will not always be the case)."

***It is always a good idea to carefully check that the number of rows returned by a join operation is what you expected. In particular, you should carefully check for rows in one table that matched to more than one row in the other table***

### Left Join

A left join includes all rows from the left dataset and the matching rows from the right dataset. If there's no match, the columns from the right dataset will be filled with NA. Here the rows of the first table are always returned, regardless of whether there is a match in the second table.

    left_joined_data <- dataset1 %>%
      left_join(dataset2, by = "common_column")

### Practice with Flights Data  

```{r left}
#Airports in the pacific time zone only 
airports_pt <- airports %>%
  filter(tz == -8) %>% 
  select(faa, name, tz)

nrow(airports_pt)
head(airports_pt)
  
#If we do inner join
nyc_dests_pt <- flights %>% 
  inner_join(airports_pt, by = c("dest" = "faa")) %>% 
  select(name, tz, dest, dep_time,sched_dep_time, arr_time, sched_arr_time, carrier)

head(nyc_dests_pt)
 
#Notice these are different.  
#key variable needs to be in BOTH data sets to get returned.  
nrow(flights)
nrow(nyc_dests_pt)
 
nyc_dests <- flights %>% 
  left_join(airports_pt, by = c("dest" = "faa"))
 
#Notice these are the same!
nrow(flights)
nrow(nyc_dests)


nyc_dests %>%
  summarize(
    num_flights = n(),
    num_flights_pt = sum(!is.na(name)),
    num_flights_not_pt = sum(is.na(name))
  )
```

### Full Join 

A full join includes all rows from both datasets. Columns from the dataset with missing values will be filled with NA where there's no match.

    full_joined_data <- dataset1 %>%
      full_join(dataset2, by = "common_column")
      
There is also a right join but this is much less common. A right join is the opposite of a left join. It includes all rows from the right dataset and the matching rows from the left dataset.    

      
### Practice with Flights Data       

```{r}
# Right Join: Keeping all airports in the pacific time zone, matching if possible
nyc_dests_right <- flights %>% 
  right_join(airports_pt, by = c("dest" = "faa"))

# Full Join: Keeping all destinations and airports, filling with NA where no match
nyc_dests_full <- flights %>% 
  full_join(airports_pt, by = c("dest" = "faa"))

# Summary of the results
summary_inner <- nyc_dests_pt %>%
  summarize(num_flights = n())

summary_left <- nyc_dests %>%
  summarize(
    num_flights = n(),
    num_flights_pt = sum(!is.na(name)),
    num_flights_not_pt = sum(is.na(name))
  )

summary_right <- nyc_dests_right %>%
  summarize(num_airports = n())

summary_full <- nyc_dests_full %>%
  summarize(num_rows = n(), 
            num_airports = sum(!is.na(name)))

# Display the summaries
summary_inner
summary_left
summary_right
summary_full

```


### Join by multiple columns

The `by` argument specifies the column(s) that should be used for matching. These join functions work well when the datasets have a shared column containing the same type of data (e.g., IDs or keys).

You can use multiple column names to define the matching conditions.

    # Syntax
    inner_join(df1, df2, by=c('x1'='y1', 'x2'='y2'))

Example Code 

```{r example_mult}
# Create emp Data Frame
emp_df=data.frame(
  emp_id=c(1,2,3,4,5,6),
  name=c("Smith","Rose","Williams","Jones","Brown","Brown"),
  superior_emp_id=c(-1,1,1,2,2,2),
  dept_id=c(10,20,10,10,40,50),
  dept_branch_id= c(101,102,101,101,104,105)
)

# Create dept Data Frame
dept_df=data.frame(
  dept_id=c(10,20,30,40),
  dept_name=c("Finance","Marketing","Sales","IT"),
  dept_branch_id= c(101,102,103,104)
)

emp_df %>% 
  inner_join(dept_df, 
        by=c('dept_id'='dept_id', 
             'dept_branch_id'='dept_branch_id'))

```


## Practice Joins

Answer the following questions either in the exercise below or in an R script

1. Which airlines had the most delayed flights departing from NYC in 2013?
2. What are the top destinations (airports) for flights departing from NYC in 2013?
3. Which airlines had the highest average departure delays from each airport?
4. What is the distribution of flight distances for flights departing from each airport?
5. Which airlines operated the most flights from each airport in NYC?


```{r join_ex, eval=FALSE}
## Q1
flights %>% 
  left_join()
## Q2

```

## Tidy Data
Tidy data is a concept popularized by Hadley Wickham, and it refers to a specific way of organizing data tables in a tabular format to facilitate data analysis. In tidy data:

* Each variable forms a column.
* Each observation forms a row.

## Pivots

Data often comes in various formats, and its structure might not be ideal for the task at hand. Pivoting helps you reorganize your data to a format that makes it easier to analyze, visualize, or model. For example, you might need to pivot data to:

* Perform Aggregations: Pivoting can simplify data aggregation and summary calculations. You can consolidate multiple columns into one and apply functions like mean, sum, etc.

* Create Visualizations: Some visualizations require specific data structures. Pivoting can help organize data to generate meaningful visualizations like time series plots, heatmaps, or faceted graphs.

* Fit Models: Certain modeling techniques might require data in a specific shape. Pivoting can help transform data into a format suitable for predictive modeling or statistical analysis.

### Pivot Longer
The `pivot_longer()` function is used to convert data from a wide format (with multiple columns) into a long format (fewer columns). It's particularly useful when you have variables spread across different columns and you want to stack them into a single column, often with corresponding values.

    pivot_longer(data, cols, names_to, values_to)

* data: The input data frame.
* cols: Columns to pivot into longer format.
* names_to: The name of the column that will store the column names from the original wide format.
* values_to: The name of the column that will store the values from the original wide format.

### Pivot Wider
The `pivot_wider()` function is used to convert data from a long format to a wide format. It's useful when you want to take distinct values from a column and spread them across new columns.

    pivot_wider(data, names_from, values_from)
    
* data: The input data frame.
* names_from: The column that contains the unique values to be transformed into column names in the wide format.
* values_from: The column that contains the values to be spread across the new wide-format columns.

```{r pivoting}
# Sample data
data_wide <- data.frame(
  country = c("USA", "Canada"),
  `2018` = c(250, 180),
  `2019` = c(260, 190),
  `2020` = c(270, 200)
)

# Using pivot_longer to reshape data
data_wide %>%
  pivot_longer(cols = -country, names_to = "year", values_to = "value")

# Sample data
data_long <- data.frame(
  country = c("USA", "Canada"),
  year = rep(c("2018", "2019", "2020"), each = 2),
  value = c(250, 260, 270, 180, 190, 200)
)
# Using pivot_wider to reshape data
data_long %>%
  pivot_wider(names_from = year, values_from = value)

```

Since we are working with a lot of datasets from the Tidyverse these datasets are already in the tidy format and thus pivoting is not needed. But once you combine dataset together there can be options to summarise new variables and thus could lead to pivoting.


## Practice Pivots

Using airline datasets from Monday, create a dataset where each airline is a row and each column is a month (1-12) to see which is the busiest months for each airline.


```{r practice_air}

```

Using the package `palmerpenguins`, pivot the dataset to that all the penguin's biological measurement (bill and flipper) are in a tidy format.

```{r penguin}

```

########################################################################
## On your own..

1. Use the `Batting`, `Pitching`, and `People` tables in the Lahman package to answer the following questions:

a) (5 points) Name every player in baseball history who has accumulated at least 300 home runs (HR) AND at least 300 stolen bases (SB). You can find the first and last name of the player in the `People` data frame. Join this to your result along with the total home runs and total bases stolen for each of these elite players.

b) (5 points) Similarly, name every pitcher in baseball history who has accumulated at least 300 wins (W) and at least 3,000 strikeouts (SO). 

c) (5 points) Identify the name and year of every player who has hit at least 50 home runs in a single season. Of those player-seasons which player had the lowest batting average (hits/at bats [H/AB])?


2. Use the nycflights13 package and the `flights` and `planes` tables to answer the following questions:

a) (5 points) What is the oldest plane (specified by the tailnum variable) that flew from New York City airports in 2013?

b) (5 points) How many airplanes are in the `flights` dataset? How many of those airplanes are included in the planes table? Is there a difference?