---
title: "Lab 1"
author: "Put Your Name Here"
date: "Due: 09-10-2023"
output: html_document
---

## RStudio Interface

The RStudio interface is composed of quadrants, each of which fulfils a unique purpose:

* The `Console` window,
* The `Source` window,
* The `Environment / History / Connections / Tutorial` window, and
* The `Files / Plots / Packages / Help / Viewer` window

Sometimes you might only see three windows and wonder where the Source window has gone in your version of RStudio. In order to use it you have to either open a file or create a new one. You can create a new file by selecting `File > New File > R Script` in the menu bar, or use the keyboard shortcut `Ctrl + Shift + N` on PC and `Cmd + Shift + N` on Mac.

#### The Console window

The console is located in the bottom-left, and it is where you often will find the output of your coding and computations. It is also possible to write code directly into the console. Let’s try the following example by calculating the sum of `10 + 5`. Click into the console with your mouse, type the calculation into your console and hit Enter/Return ↵ on your keyboard. The result should be pretty obvious:

```{r, eval=FALSE}
  # We type the below into the console
  10 + 5
#> [1] 15
```



You just successfully performed your first successful computation. I know, this is not quite impressive just yet. R is undoubtedly more than just a giant calculator.

In the top right of the console, you find a symbol that looks like a broom. This one is quite an important one because it clears your console. Sometimes the console can become very cluttered and difficult to read. If you want to remove whatever you computed, you can click the broom icon and clear the console of all text. I use it so frequently that I strongly recommend learning the keyboard shortcut, which is `Ctrl + L` on PC and Mac.

#### The Source window
In the top left, you can find the source window. The term ‘source’ can be understood as any type of file, e.g. data, programming code, notes, etc. The source panel can fulfil many functions, such as:

* Inspect data in an Excel-like format 

* Open programming code, e.g. an R Script 

* Open other text-based file formats, e.g.

  - Plain text (.txt),

  - Markdown (.rmd),

  - Websites (.html),

  - LaTeX (.tex),

  - BibTex (.bib),

* Edit scripts with code in it,

* Run the analysis you have written.

In other words, the source window will show you whatever file you are interested in, as long as RStudio can read it - and no, Microsoft Office Documents are not supported. Another limitation of the source window is that it can only show text-based files. So opening images, etc. would not work.

#### The Environment / History / Connections / Tutorial window
The window in the top right shows multiples panes. The first pane is called Environment and shows you objects which are available for computation. One of the first objects you will create is your dataset because, without data, we cannot perform any analysis. Another object could be a plot showing the number of male and female participants in your study. Besides datasets and plots, you will also find other objects here, e.g. lists, vectors and functions you created yourself. Don’t worry if none of these words makes sense at this point. We will cover each of them in the upcoming chapters. For now, remember this is a place where you can find different objects you created.

#### The Files / Plots / Packages / Help / Viewer window
The last window consists of five essential panes. The first one is the Files pane. As the name indicates, it lists all the files and folders in your root directory. A root directory is the default directory where RStudio saves your files, for example, your analysis. Thus, the Files pane is an easy way to load data into RStudio and create folders to keep your research project well organised.

Since the Console cannot reproduce data visualisations, RStudio offers a way to do this very easily. It is through the Plots pane. This pane is exclusively designed to show you any plots you have created using R. 

If you wish to delete the plot, you can click on the red circle with a white x symbol. This will delete the currently visible plot. If you wish to remove all plots from this pane, you can use the broom. There is also an option to export your plot and move back and forth between different plots.

The next pane is called Packages. Packages are additional tools you can import and use when performing your analysis. A frequent analogy people use to explain packages is your phone and the apps you install. Each package you download is equivalent to an app on your phone. It can enhance different aspects of working in R, such as creating animated plots, using unique machine learning algorithms, or simply making your life easier by doing multiple computations with just one single line of code.

If you are in dire need of help, RStudio provides you with a Help pane. You can search for specific topics, for example how certain computations work. The Help pane also has documentation on different datasets that are included in R, RStudio or R packages you have installed. If you want a more comprehensive overview of how you can find help, have a look at CRAN’s [‘Getting Help with R’](https://www.r-project.org/help.html) webpage.

#### Customise your user interface
As a last remark in this chapter, I would like to make you aware that you can modify each window. There are three basic adjustments you can make:

* Hide panes by clicking on the window symbol in the top right corner of each window,

* Resize panes by dragging the border of a window horizontally or vertically, or

* Add and remove panes by going to `RStudio > Preferences > Pane Layout`, or use the keyboard shortcut `⌘ + ,` if you are on a Mac. Unfortunately, there is no default shortcut for PC users.

If you want a fully customised experience you can also alter the colour scheme of RStudio itself (`RStudio > Preferences > Appearance`).


## Install Packages

Many of R’s most useful functions do not come preloaded when you start R, but reside in packages that can be installed on top of R. R packages are similar to libraries in C, C++, and Javascript, packages in Python, and gems in Ruby. An R package bundles together useful functions, help files, and data sets. You can use these functions within your own R code once you load the package they live in. Usually the contents of an R package are all related to a single type of task, which the package helps solve. R packages will let you take advantage of R’s most useful features: its large community of package writers (many of whom are active data scientists) and its prewritten routines for handling many common (and exotic) data-science tasks.


> You may hear R users (or me) refer to “base R.” What is base R? It is just the collection of R functions that gets loaded every time you start R. These functions provide the basics of the language, and you don’t have to load a package before you can use them.

### How to install R packages

To use an R package, you must first install it on your computer and then load it in your current R session. The easiest way to install an R package is with the install.packages R function. Open R and type the following into the command line:

> install.packages("<package name>")
  
This will search for the specified package in the collection of packages hosted on the CRAN site. When R finds the package, it will download it into a libraries folder on your computer. R can access the package here in future R sessions without reinstalling it. Anyone can write an R package and disseminate it as they like; however, almost all R packages are published through the CRAN website. CRAN tests each R package before publishing it. This doesn’t eliminate every bug inside a package, but it does mean that you can trust a package on CRAN to run in the current version of R on your OS.

You can install multiple packages at once by linking their names with R’s concatenate function, c. For example, to install the tidyverse, run:

> install.packages("tidyverse")
  
If this is your first time installing a package, R will prompt you to choose an online mirror of to install from. Mirrors are listed by location. Your downloads should be quickest if you select a mirror that is close to you. If you want to download a new package, try the Austria mirror first. This is the main CRAN repository, and new packages can sometimes take a couple of days to make it around to all of the other mirrors.


#### Loading Packages

Installing a package doesn’t immediately place its functions at your fingertips. It just places them on your computer. To use an R package, you next have to load it in your R session with the command:

> library(package name)
  
Notice that the quotation marks have disappeared. You can use them if you like, but quotation marks are optional for the library command. (This is not true for the install.packages command).

library will make all of the package’s functions, data sets, and help files available to you until you close your current R session. The next time you begin an R session, you’ll have to reload the package with library if you want to use it, but you won’t have to reinstall it. You only have to install each package once. After that, a copy of the package will live in your R library. To see which packages you currently have in your R library, run:

> library()
  
library() also shows the path to your actual R library, which is the folder that contains your R packages. You may notice many packages that you don’t remember installing. This is because R automatically downloads a set of useful packages when you first install R.

## File Types

There are many types of files that you will interact with when using R and Rstudio.

1. R Files
* .R - an R script
* .Rmd - an RMarkdown script - incorporates code and texts together to *knit* a document together. Knitting an rmd file compiles the text and code together to create another file such as an .html or .pdf This file is an Rmd file and a code chunk is created with a shortcut Ctrl + Alt + I (Windows/Linux) or Cmd + Option + I (macOS). 
* .Rproj - an R Project file which works with Rstudio to help you organize your different coding projects. The working directory will automatically be set to the directory that the .RProj file is located in allowing easier organization and access to your data and files.
2. Data Files
* .csv - comma seperated values file
* .xlxs - excel file
* .txt - text file


## Basics of R

### Assignment

We assign the value of an object on the left hand side, to the name (expression) on the right hand side. This tells R what to call the object or variable.

```{r assign}
# Preferred syntax
height <- 442

# These also work
442 -> height
height = 442

height
```


### Comments
Comments are notes (or comments) that are made in a code chunk that does not get run but is there for human use.
```{r comment}
# This is how you make a comment!
```



> R is Case Sensitive

```{r note}
# These are all different objects
name <- "greg" 
Name <- "steve"
NAME <- "mary"
```

### Operations
R is a big calculator!
```{r calc}
# Add
3 + 3

# Subtract
6 - 3

# Multiply
4 * 3

# Divide
6 / 2

# Floored quotient
floor(7 / 2)

# Ceiling quotient
ceiling(7 / 2)

# Remainder or modulus
7 %% 2

# Absolute value
abs(-5)

```



## Classes and Data Structures 

### Classes

A variable can store different types of values such as numbers, characters, etc. These different types of data that we can use in our code are called data types which belong to a certain class.

Common data types and classes 

1. numeric: any real number(s)
2. character: strings or individual characters, quoted
3. integer: any integer(s)/whole numbers
4. factor: categorical/qualitative variables
5. logical: variables composed of TRUE or FALSE
6. Date/POSIXct: represents calendar dates and times


```{r types}
# Default to numeric
height <- 442
class(as.integer(height))

height <- 442.0
class(height)

# Character
height <- "really tall"
class(height)

# Can force it to be an integer
height <- as.integer(442)
class(height)
object.size(height)
```

Convert to different types

```{r convert}
# Convert numeric to integer
a <- 5
a <- as.integer(a)

# Convert character to numeric
a <- "5"
a <- as.numeric(a)

# Convert numeric to character
a <- 5
a <- as.character(a)
```

### Data Structures

#### Vectors in R
Vectors in R can only contain elements of a single class. Here's an example of creating and working with a numeric vector:

```{r vector}
x <- c(4, 1, 3, 8, 6, 7, 5, 3, 0, 9)
x
```

##### Subsetting Vectors
Subsetting allows you to extract specific elements from a vector:

```{r subset_vector}
x[1] ## first element
x[1:5] ## first 5 using the colon
```

##### Vector Operations
You can perform various operations on vectors:

```{r vector_opt}
y <- 1:10
x + y
1 / x
```


#### Factors

Factors are essential in statistics. They represent categorical variables:

```{r factor}
d <- factor(c("A", "A", "B", "C", "A", "B"))
d
```

#### The `paste` Function
The `paste` function is used to combine character strings:

```{r paste}
f <- paste("x", 1:5, sep = "")
f
```


#### Matrices in R
Matrices are two-dimensional arrays in R:
```{r matrix}
mat <- matrix(c(1:9), ncol = 3, byrow = TRUE)
mat
```

##### Subsetting Matrices
Subsetting matrices involves specifying both row and column indexes:
```{r subset_matrix}
mat[1, 1]
mat[1, ]
```

#### Lists in R
Lists are versatile R objects, able to hold elements of various classes:

```{r list}
l <- list(3, rep(0, 3), matrix(c(1:4), ncol = 2))
l
```


##### Indexing Lists
Lists can be indexed using double brackets or the `$` notation:

```{r index_list}
l[[3]]
l$three
```


#### Data Frames in R
Data frames combine properties of matrices and lists, suitable for data analysis:
```{r dataframe}
df <- data.frame(V1 = 1:10, V2 = rep(1, 10), V3 = seq(1, 20, 2))
df
```


##### Subsetting Data Frames
Data frames can be subsetted by index or name:

```{r subset_dataframe}
df[,1]
df[1,]
```


#### Summary and Table Functions
Use `summary` for a five-number summary and `table` for frequency tables:

```{r sum_tab}
summary(df)
table(df$V3)
```

## Git and Github

### Git
Git is a version control system that helps you keep track of changes in your code or any other files. It's like a time machine for your work. Git allows you to:

* Track Changes: You can record every change you make to your files. This includes adding new files, editing existing ones, and deleting files.

* Collaborate with Others: Git is designed for collaboration. Multiple people can work on the same project simultaneously, and Git helps to manage changes and merge them seamlessly (ish).

* Create Backups: Git acts as a backup system. If something goes wrong, you can easily restore your project to a previous state.

* Work Offline: You don't need an internet connection to use Git. You can commit changes to your local repository, and when you're online, you can sync those changes with a remote repository (like GitHub).

### GitHub
GitHub is a web-based platform that uses Git for version control. It's like a social network for programmers and a hosting platform for Git repositories. Here's what you can do with GitHub:

* Store Code in the Cloud: You can store your Git repositories online. This is particularly useful because it acts as a backup, and you can access your code from anywhere.

Collaborate with Others: GitHub is an excellent platform for team collaboration. You can invite team members to your project, and everyone can work on the same codebase. This is what we will be using for our class to acess notes and labs.

* Share Your Work: GitHub allows you to share your code with the world. You can make your repositories public so that anyone can see your code and contribute to it, or you can keep them private for restricted access.

* Issue Tracking: GitHub provides tools for issue tracking and project management. You can use it to keep track of tasks, bugs, and feature requests.


Install Git: Install Git on your computer. Git can be downloaded from the [official website](https://git-scm.com/downloads).

Set Up Git: Open a terminal (or Git Bash on Windows), note that the terminal is present in the console, and configure Git with their name and email address using the following commands:

    git config --global user.name "Your Name"
    git config --global user.email "youremail@example.com"
    

____________________________________________________________________
## On your own..

1. I want you to connect your Rstudio to the class Github page [https://github.com/menawhalen/STAT_401](https://github.com/menawhalen/STAT_401).

* Clone the Repository: In RStudio, go to "File" > "New Project" > "Version Control" > "Git." Paste the URL of your GitHub repository (https://github.com/menawhalen/STAT_401) and choose a directory to clone it into. This directory will be on your computer locally so organize it accordingly.

* Pull the Latest Changes: Before working on any labs, pull in the terminal to get the latest version of the repository. The pull is the green arrow under the Git tab. You will not be committing or pulling to this class. Just *pull*.

2. Create a numeric vector containing the values 5, 10, 15, 20, and 25. Calculate the mean of this vector and assign it to a variable called `mean_value`. Print the `mean_value`

```{r ex2}

```


3. Create a data frame with the following information about three of your friends:

* Name (character)
* Age (numeric)
* Favorite Color (character)

Make sure to include at least three rows of data. Print the entire data frame. Name it whatever you like.

```{r ex3}

```

4. Create two numeric vectors: `vector1` with values 1, 2, 3, and `vector2` with values 4, 5, 6. Perform the following operations:

* Add `vector1` and `vector2` and store the result in a new vector called `sum_vector.`
* Subtract `vector2` from `vector1` and store the result in a new vector called `diff_vector.`
* Multiply `vector1` and `vector2` element-wise and store the result in a new vector called `prod_vector.`
* Divide `vector1` by `vector2` element-wise and store the result in a new vector called `quot_vector.`

Print all four result vectors.

```{r ex4}

```

5. Knit this document and submit the Rmd file and the html version to Sakai under Lab 1. Both Rmd and html files need to be submitted to be graded.